---
title: "Investigating COVID-19"
author: "Andrew Lujan"
date: '2022-05-04'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project will analyze COVID-19 data from [Kaggle](https://www.kaggle.com/datasets/lin0li/covid19testing). 

The main purpose of our analysis is to answer the following question:
* **Which countries have had the highest number of positive cases against the number of tests**?

## Understanding the Data

### Load the dataset from the `covid19.csv` file for a quick exploration

```{r}
library(readr)

## Loading the dataset
covid_df <- read_csv("covid19.csv")

```

We successfully pulled in the data using the `readr()` function. 

```{r}

## Check the dimensions of the dataset

dim(covid_df)

```

```{r}

## Get the names of the columns and find out what they represent

vector_cols <- colnames(covid_df)

## Displaying the variable vector_cols

vector_cols

## Showing the first rows of the dataset

head(covid_df)

## Showing a global view of the dataset

library(tibble)

glimpse(covid_df)

```
We have various vector column names listed below:
* Date, name of the continent, country code, country regions, province or state, positive, hospitalized, recovered, death, total people tested, active, hospitalized currently, daily test totals, daily postive rates

### Dimensions of the dataset

The dataset contains `14` columns and `10,903` rows. It provides information on total cases (per day and cumulatively) of COVID-19 positive cases, deaths, tests performed, and hospitalizations for each country through the column's names stored in the variable `vector_cols`. 

1. The variable `vector_cols` contains a character vector. 
2. The glimpse function is particulary useful because it lists the names of the columns, the dimension of the table, column types, and can replace the other functions we've used already. 
## Isolating rows we need
Looking at the data, we can see that the column `Province_State` column has mixture of data from different levels. We need to filter the data so our analysis will not be biased. 

1. We'll filter rows related to `All_States` from the `Province_State` column and then remove that column from the covid_df

```{r}

## Filter rows related to All_States from the `Province_State`

library(dplyr)

# Filter
covid_df_all_states <- covid_df %>%
  filter(Province_State == "All States") %>%
  select(-Province_State)


```

We are able to remove `Province_State` without losing information because after filtering this column only contains the values "All_States".

### Isolating the columns we need

- Create a dataset for the daily columns from `covid_df_all_states` dataframe.

The description of the dataset's columns are below:

Let's recall the description of the dataset's columns.

1. `Date`: Date
2. `Continent_Name`: Continent names
3. `Two_Letter_Country_Code`: Country codes
4. `Country_Region`: Country names
5. `Province_State`: States/province names; value is `All States` when state/provincial level data is not available
6. `positive`: Cumulative number of positive cases reported.
7. `active`: Number of actively cases on that **day**.
8. `hospitalized`: Cumulative number of hospitalized cases reported.
9. `hospitalizedCurr`: Number of actively hospitalized cases on that **day**.
10. `recovered`: Cumulative number of recovered cases reported.
11. `death`: Cumulative number of deaths reported.
12. `total_tested`: Cumulative number of tests conducted.
13. `daily_tested`: Number of tests conducted on the **day**; if daily data is unavailable, daily tested is averaged across number of days in between.
14. `daily_positive`: Number of positive cases reported on the **day**; if daily data is unavailable, daily positive is averaged across number of days in.


We are planning on working with mainly daily data, so we will extract the columns that are related to the daily measures. 

```{r}

## Selecting columns with daily numbers

covid_df_all_states_daily <- covid_df_all_states %>%
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)

head(covid_df_all_states_daily) 

```

## Extracting the Top Ten countries in the number of tested cases

- How can we get the overall number of COVID-19 tested, positive, active and hospitalized cases by country since we currently have daily data?
  * group_by(), summarize()

- How do we then extract the top ten?
  * arrange() by top 10 head()
  
```{r}

## Summarize dataframe by computing sum of daily totals and group by the Country_Region column

covid_df_all_states_daily_sum <- covid_df_all_states_daily %>%
  group_by(Country_Region) %>%
  summarize(
    tested = sum(daily_tested),
    positive = sum(daily_positive),
    active = sum(active),
    hospitalized = sum(hospitalizedCurr)
    ) %>%
      arrange(desc(tested))

covid_df_all_states_daily_sum

```

```{r}

# Extracting the top 10 rows


covid_top_10 <- head(covid_df_all_states_daily_sum, 10)

covid_top_10

```

  
## Which countries have had the highest number of positive cases against the number of tests

### Creating vectors from the covid_top_10 dataframe for analysis

```{r}

countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized



```


### Naming the vectors

```{r}

names(positive_cases) <- countries
names(tested_cases) <- countries
names(active_cases) <- countries
names(hospitalized_cases) <- countries



```

### Identify the top three positive against tested cases

```{r}

# Finding the top 3 positive against tested cases
positive_cases
sum(positive_cases)
mean(positive_cases)

```

```{r}

positive_cases / tested_cases

```


### Storing the top 3 in a vector

```{r}

positive_tested_top_3 <- c("United Kingdom" = .11, "United States" = .10, "Turkey" = .08)

```

## Keeping relvant information

```{r}

## Creating vectors for the top 3
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)

## Creating a matrix that combines this information
covid_mat <- rbind(united_kingdom, united_states, turkey)

# Renaming the columns using the colnames() function
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", "hospitalized")

# Displaying the matrix
covid_mat

```

Now that we have the top 3 countries with the highest number of positive COVID-19 cases, we are going to move the final step which is circling back to our questions and answering them. 
### Answering the research questions

```{r}

question <- "Which countries have had the highest number of positive cases against the number of tests?"

answer <- c("Positive tested cases" = positive_tested_top_3)

# Datasets list
datasets <- list(
  original = covid_df, 
  allstates = covid_df_all_states,
  daily = covid_df_all_states_daily,
  top_10 = covid_top_10
)

# Matrices list

matrices <- list(covid_mat)

# Vectors list

vectors <- list(vector_cols, countries)

data_structure_list <- list("dataframe" = datasets, "matrix" = matrices, "vector" = vectors)

# Creating the Covid Analysis List

covid_analysis_list <- list(question, answer, data_structure_list)

covid_analysis_list[[2]]

```

